Concurrent Programming with CXL

### 2020-06-22 book/paper.tex
Latest

A good example of a resource with multiple units is the
so-called \emph{bounded buffer} (aka \emph{ring buffer}).
\index{ring buffer}
Suppose you have a \emph{producer} (some process) wanting to
communicate a stream of items with a \emph{consumer} (another process).
A bounded buffer is essentially
a queue implemented using a circular buffer
\index{circular buffer}
of a certain length and two pointers:
one where the producer inserts items, and one where the consumer extracts
items.  If the
buffer is full, the producer must wait; if the buffer is empty, the
consumer must wait.
This problem is known as the ``Producer/Consumer Problem'' and was
proposed by Dijkstra~\cite{EWD329}.
Here we generalize it to allowing multiple producers and multiple consumers
all sharing the same bounded buffer.

Figure~\ref{fig:pc} gives a high-level description of what we want.  The
two numbers in a state specify
the number of items that the producers still can produce and
the number of items that the consumers have consumed.  In this case
there are three items to produce initially.

@@

Figure~\ref{fig:boundedbuffer} presents the implementation of a bounded
buffer using semaphores.  It features a circular bounded buffer \texttt{buf} with
slots numbered 1 through \texttt{NSLOTS}.  There are two indexes into
the buffer: \texttt{b\_in} specifies where the next produced item is inserted,
while \texttt{b\_out} specifies from where the next can be consumed.
As there may be multiple producers and multiple
consumers, updates to the indexes, which are shared variables, must be protected.
To this end we use two semaphores as locks: \texttt{l\_in} for \texttt{b\_in}
and \texttt{l\_out} for \texttt{b\_out}.

Without additional synchronization, the indexes may overflow and point to invalid
entries in the circular buffer.
To this end there is a semaphore \texttt{n\_full} that
keeps track of the number of filled entries in the buffer and a semaphore
\texttt{n\_empty} that keeps track of the number of empty entries in the buffer.

To add an item to the bounded buffer (\texttt{produce(item)}), the producing
process first has to wait until there is room in the bounded buffer.
To this end, it invokes \texttt{P(n\_empty)}, which waits until
\texttt{n\_empty > 0} and atomically decrements \texttt{n\_empty} once this
is the case.  In other words, the producer procures an empty slot.
Next, it adds the item to the next position in the buffer.
Since there may be more than one empty slot, multiple producers may attempt
to do the same thing concurrently.  Hence the use of the \texttt{l\_in}
semaphore used as a lock.  Finally, once the item is added, the process
increments the \texttt{n\_full} semaphore to signal to consumers that
an item is available.

The consumer code is symmetric.

Note the quite different usage of the \texttt{l\_} semaphores and \texttt{n\_}
semaphores.
People often call semaphores that are used as locks \emph{binary semaphores},
\index{binary semaphore}
as their counters only take on the values 0 and 1.
They protect critical sections.
The second type of semaphores are called \emph{counting semaphores}.
\index{counting semaphore}
They can be used to send signals between processes.
However, binary and counting semaphores are implemented the same way.

We can run the code through CXL, but what does it test?  There are no
assert statements in the code.
We can split the functionality of this code into two pieces.  One is
the synchronization part: producers should never run more than \texttt{NSLOTS}
ahead of the consumers, but they should be able to produce up to \texttt{NSLOTS}
items even if there are no consumers.  The other part is the data: we want
to make sure that the items are sent through the buffer in order and without
loss.  We will first focus on the synchronization.

The first part can be tested with the code given in Figure~\ref{fig:pc}, by
experimenting with different non-negative values for the three constants.
In particular, any run with a positive \texttt{NSLOTS} and
$\mathtt{NCONSS} \le \mathtt{NPRODS} \le \mathtt{NCONSS} + \mathtt{NSLOTS}$
should terminate, while other values should lead to processes blocking.
One could write a script like the one in Figure~\ref{fig:pcscript}.

@@

\section*{Exercises}
\begin{enumerate}
\item Change the code in Figure~\ref{fig:boundedbuffer} so that
\texttt{produce()} fails if the buffer is full rather than block.
\texttt{produce()} should return \texttt{False} if it fails
and return \texttt{True} if it succeeds.
\item Implement a \emph{bounded stack}.  Method \texttt{push(item)}
should block if the stack is full.  Method \texttt{pop()} should block
until the stack is non-empty and then return an item.
\item Implement a GPU allocator.  There are \texttt{N} GPUs identified
by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease(gpu)} releases the given GPU.  It never needs
to block.
\end{enumerate}

@@

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections}
\index{conditional critical section}
(CCSs)~\cite{Hoare73}.
\glsadd{conditional critical section}
All critical sections are, in a way, conditional.  Even the basic critical section
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

@@

Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
problem, there are two: a process that wants to read may have to wait for a
writer to leave the critical regiion, while a process that wants to write may
have to wait until all readers have left the critical section.  If there are $N$
waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
each waiting condition and an extra one that processes use when they first try
to enter the critical section and do not yet know if they need to wait.
We shall call this the \texttt{mutex}.

An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
must always be 0 or 1, and it should be 0 when a process is accessing the shared
variables.
As it were, these semaphores taken together are emulating a single binary semaphore.
Initially, \texttt{mutex} is~1 and the other $N$ semaphores
are all~0.  To maintain the invariant, each process alternates \texttt{P} and
\texttt{V} operations, starting with \texttt{P~\&(mutex)} when it tries to
enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
(yet to be determined).

@@

Either way, it vacates one of the semaphores next.
\texttt{V\_one()} is the baton passing function that selects which of the
binary semaphores to vacate.
Method \texttt{V\_one()} first checks to see if there are any readers or
writers waiting.  If there are readers waiting and there are no writers
in the critical section, it vacates the semaphore associated with
readers waiting.  If there are writers waiing and there are no readers
nor writers in the critical section, it vacates the semaphore associated
with writers waiting.  If both conditions hold, it could vacates either
one---it would not matter.  However, it can only vacate one of the semaphores.
If none of the conditions hold, \texttt{V\_one()} vacates \texttt{mutex}.

Let us return now to \texttt{rlock\_acquire}, in the case that the process
found there is a writer in the critical section, it vacates
\texttt{mutex} and procures \texttt{r\_sema},
blocking on the semaphore associated
with waiting readers.  When later some other process passes the baton to
it by incrementing \texttt{r\_sema}), it updates the state by decrementing
\texttt{r\_waiting} and incrementing \texttt{r\_entered}.  It
then invokes \texttt{V\_one()} one more time.

>>>>

A good example of a resource with multiple units is the
so-called \emph{bounded buffer} (aka \emph{ring buffer}).
\index{ring buffer}
Suppose you have a \emph{producer} (some process) wanting to
communicate a stream of items with a \emph{consumer} (another process).
A bounded buffer is essentially
a queue implemented using a circular buffer
\index{circular buffer}
of a certain length and two pointers:
one where the producer inserts items, and one where the consumer extracts
items.  If the
buffer is full, the producer must wait; if the buffer is empty, the
consumer must wait.
This problem is known as the ``Producer/Consumer Problem'' and was
proposed by Dijkstra~\cite{EWD329}.

The producer/consumer pattern is common.  Processes may be arranged
in pipelines, where each upstream process is a producer and each downstream
process is a consumer.
Or processes may be arranged in a manager/worker pattern, with a manager
producing jobs and workers executing them in parallel.

In general, we allow multiple producers and multiple consumers
all sharing the same bounded buffer.
Figure~\ref{fig:pc} gives a high-level description.  The
two numbers in a state specify
the number of items that the producers still can produce and
the number of items that the consumers have consumed.  In this case
there are three items to produce initially.

@@

Figure~\ref{fig:boundedbuffer} presents the implementation of a bounded
buffer using semaphores.  It features a circular bounded buffer \texttt{buf} with
slots numbered 1 through \texttt{NSLOTS}.  There are two indexes into
the buffer: \texttt{b\_in} specifies where the next produced item is inserted,
while \texttt{b\_out} specifies from where the next can be consumed.
As there may be multiple producers and multiple
consumers, updates to the indexes, which are shared variables, must be protected.
To this end, we use two semaphores as locks: \texttt{l\_in} for \texttt{b\_in}
and \texttt{l\_out} for \texttt{b\_out}.

Without additional synchronization, the indexes may overflow and point to invalid
entries in the circular buffer.
For this, there is a semaphore \texttt{n\_full} that
keeps track of the number of filled entries in the buffer and a semaphore
\texttt{n\_empty} that keeps track of the number of empty entries in the buffer.

To add an item to the bounded buffer (\texttt{produce(item)}), the producing
process first has to wait until there is room in the bounded buffer.
To this end, it invokes \texttt{P(n\_empty)}, which waits until
\texttt{n\_empty > 0} and atomically decrements \texttt{n\_empty} once this
is the case.  In other words, the producer procures an empty slot.
Next, it adds the item to the next position in the buffer.
Since there may be more than one empty slot, multiple producers may attempt
to do the same thing concurrently.  Hence the use of the \texttt{l\_in}
semaphore used as a lock.  Finally, once the item is added, the process
increments the \texttt{n\_full} semaphore to signal to consumers that
an item is available.
The consumer code is symmetric.

Note the quite different usage of the \texttt{l\_} semaphores and \texttt{n\_}
semaphores.
People often call semaphores that are used as locks \emph{binary semaphores},
\index{binary semaphore}
as they only take on the values 0 and 1.
They protect critical sections.
The second type of semaphores are called \emph{counting semaphores}.
\index{counting semaphore}
They can be used to send signals between processes.
However, binary and counting semaphores are implemented the same way.

We can run the code through CXL, but what would it test?  There are no
assert statements in the code.
We can split the functionality of this code into two pieces.  One is
the synchronization part: producers should never run more than \texttt{NSLOTS}
ahead of the consumers, but they should be able to produce up to \texttt{NSLOTS}
items even if there are no consumers.  The other part is the data: we want
to make sure that the items are sent through the buffer in order and without
loss.  We will first focus on the synchronization.

The first part can be tested with the code given in Figure~\ref{fig:pc}, by
experimenting with different non-negative values for the three constants.
In particular, any run with a positive \texttt{NSLOTS} and
$\mathtt{NCONSS} \le \mathtt{NPRODS} \le \mathtt{NCONSS} + \mathtt{NSLOTS}$
should terminate, while other values should lead to processes blocking.
To this this, one can write a script like the one in Figure~\ref{fig:pcscript}.

@@

\section*{Exercises}
\begin{enumerate}
\item Change the code in Figure~\ref{fig:boundedbuffer} so that
\texttt{produce()} fails if the buffer is full rather than block.
\texttt{produce()} should return \texttt{False} if it fails
and return \texttt{True} if it succeeds.
\item Implement a \emph{bounded stack}.  Method \texttt{push(item)}
should block if the stack is full.  Method \texttt{pop()} should block
until the stack is non-empty and then return an item.
\item Implement a thread-safe GPU allocator.  There are \texttt{N} GPUs identified
by the numbers
1 through \texttt{N}.  Method \texttt{gpuAlloc()} returns the identifier
of an available GPU, blocking if there is currently no available GPU.
Method \texttt{gpuRelease(gpu)} releases the given GPU.  It never needs
to block.
\end{enumerate}

@@

Reader/writer locks and bounded buffers are both examples of what are
sometimes called \emph{conditional critical sections}
\index{conditional critical section}
(CCSs)~\cite{Hoare73}.
\glsadd{conditional critical section}
All critical sections are, in a way, conditional.  Even the basic critical section
has the condition that a process can only enter if no other process
is in the critical section.  Reader/writer locks refined that condition, loosening
the restriction for improved performance.
In the case of bounded buffers, there were two CCSs.  First, a CCS
to add an entry to the bounded buffer that can only be entered by one
process and only if there is room in the bounded buffer (\texttt{produce}).
Second, a CCS that can only be entered by one process and
only if there is an item in the bounded buffer (\texttt{consume}).

@@

Step 1 is to enumerate all waiting conditions.  In the case of the reader/writer
problem, there are two: a process that wants to read may have to wait for a
writer to leave the critical region, while a process that wants to write may
have to wait until all readers have left the critical section.  If there are $N$
waiting conditions, the SBS technique will use $N+1$ binary semaphores: one for
each waiting condition and an extra one that processes use when they first try
to enter the critical section and do not yet know if they need to wait.
We shall call this the \texttt{mutex}.

An important invariant of the SBS technique is that the sum of the $N+1$ semaphores
must always be 0 or 1, and it should be 0 when a process is accessing the shared
variables and 1 when not.
As it were, these semaphores taken together are emulating a single binary semaphore.
Initially, \texttt{mutex} is~1 and the other $N$ semaphores
are all~0.  To maintain the invariant, each process alternates \texttt{P} and
\texttt{V} operations, starting with \texttt{P(\&mutex)} when it tries to
enter the CCS, and ending with a \texttt{V} operation on one of the semaphores
(yet to be determined).

@@

Either way, it vacates one of the semaphores next.
\texttt{V\_one()} is the baton passing function that selects which of the
binary semaphores to vacate.
Method \texttt{V\_one()} first checks to see if there are any readers or
writers waiting.  If there are readers waiting and there are no writers
in the critical section, it vacates the semaphore associated with
readers waiting.  If there are writers waiing and there are no readers
nor writers in the critical section, it vacates the semaphore associated
with writers waiting.  If both conditions hold, it could vacate either
one---it would not matter.  However, it can only vacate one of the semaphores.
If none of the conditions hold, \texttt{V\_one()} vacates \texttt{mutex}.

Let us return now to \texttt{rlock\_acquire}. In the case that the process,
say $p$,
found there is a writer in the critical section, it vacates
\texttt{mutex} and procures \texttt{r\_sema},
blocking on the semaphore associated
with waiting readers.  When later some other process passes the baton to
$p$ by incrementing \texttt{r\_sema}, $p$ updates the state by decrementing
\texttt{r\_waiting} and incrementing \texttt{r\_entered}.  $p$
then invokes \texttt{V\_one()} one more time.
